Changes in accessibility to emergency and community food services during COVID-19 and implications for low income populations in Hamilton, Ontario

### 2021-08-27 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.Rmd

Food insecurity is a significant issue for many Canadian households and is associated with reductions in nutrition and poor mental and physical health outcomes. While food bank use can provide some relief, the levels of need in the City of Hamilton were on the rise before the onset of the COVID-19 pandemic, with a troubling increase in the number of visits made by children. While temporary financial relief measures may have helped reduce food bank use in the summer of 2020, the pandemic has in all probability increased food insecurity for many households. To compound matters, the pandemic has also resulted in major disruptions, including to employment, mobility alternatives, and to food bank and food service delivery. In response, this research has sought to better understand accessibility to food banks and food service locations, as well as how the closure of some locations over the pre- and during-COVID time periods affected the potential for low-income households to reach these amenities. Using the balanced floating catchment area method to account for population demand and congestion effects at service points, we estimated multi-modal accessibility to food bank and food service locations for low-income households. The weighting of travel time estimates by the modal split in different zonal geographies tailors the results to patterns of travel behaviour captured in the regional travel survey. Moreover, our parcel-level analysis presents a disaggregate approach to estimating accessibility based on the locations of residential parcels and dwellings.

Our results show that while accessibility levels were lower in the city's more car-oriented suburban and rural areas to begin with, the closure of 14% of the city's food bank and low-cost or free food service locations during the pandemic resulted in an overall decrease in accessibility across the city. However, these effects were not uniform over space or for different population groups. Since the balanced floating catchment area method takes into account changes in demand and congestion for service providers, the closure of food bank and food service locations can reverberate throughout the catchment areas of the whole city. For some suburban zones, the closure of a relatively high level-of-service location results in the remaining services being spread over a larger population. In others, high auto dependence for trips leads to decreases in accessibility that accumulate due to the loss of several food bank and service locations initially reachable within 15 minutes by car. Reductions in accessibility in the city's more urban north end, where low-income households conduct higher proportions of trips by transit and walking, emphasize the importance of geographic proximity in the potential to reach food bank and service locations for these residents. Beyond geography, the results also highlight the differential impact of food banks closing during COVID-19 on population groups, with seniors and children most affected.

In the absence of information regarding how food insecure households travel to food banks and related services, we examined accessibility to food banks using a 15 minutes (one-way) travel time threshold. The fact that we must rely on a standard created for accessible drinking water in the developing world only serves to highlight the tragedy of food insecurity in an affluent country like Canada. More broadly, it points to the absurd need to understand how a bad situation was made worse by the pandemic: in effect, the analysis reveals that disparities in accessibility to food banks and services predated the pandemic, that the pandemic contributed to the deterioration of these services, and that populations already in distress, particularly children, ended up in an even more adverse state. How much worse, it is impossible to say, mainly because there is also a dearth of information, let alone standards, regarding what is an *acceptable* or *sufficient* level of service when it comes to food banks. Whether this is one facility per 1,000 population or three facilities per 1,000 population, or any other number, we know not. But like international human rights standards for accessibility to clean drinking water within 30 minutes (round-trip), this work strongly suggests that adopting better defined standards for access to food banks that move beyond generalized conceptualizations of "access" to consider the geographic component of accessibility (and its interactions with travel behaviour and individual/household socioeconomic and demographic traits) would make a meaningful contribution to understanding food access and "food bank deserts" for food insecure populations.

In summary, while previous work has noted the important role of geography alongside other socio-economic and demographic indicators in household access to healthy food, this is, to the best of our knowledge, the first study that has focused on the geographic component of food bank accessibility. Beyond accounting for the inflation of demand that occurs in traditional floating catchment area methods, our application of the balanced floating catchment area approach offers a novel analysis of food bank and food service accessibility that is sensitive to the locations of low-income households, information on their age distributions and typical trip-making behaviour, food bank and food service locations and their operations over time, and the characteristics of the city's multi-modal transportation network. Despite the limitations noted above, the findings from this analysis can provide quantitative evidence to help urban planners and community administrators strategically allocate food bank services considering accessibility for different population groups, especially when considering the unequal distribution of impacts from the COVID-19 pandemic.

In terms of future research, it is important to note that the degree to which our low-income cut-off of CAD\$40,000 reflects food insecurity in the different zones of the study area is not known. We also consider all food banks and food service locations equally. Information on differential capacities at food provider locations is currently not collected, and given the closure of facilities was not possible to obtain. Data on services, such as the number of meals served, could be used to refine the analysis in the future. Moreover, while the travel survey allows us to model multi-modal accessibilities that align with travel behaviour observed in the travel survey and capture differences in accessibility by age categories, the use of travel survey data for modelling food insecurity also has its limitations. Research into the population weighting methods used in the TTS note that the survey may under-count the lowest- and highest-income households in the survey study region, although the magnitude of this under-counting is unknown, and approximately 20% of respondents to the 2016 survey did not report their income [@rose2018transportation]. In that regard, the modal splits of low-income households observed through the TTS data may not accurately reflect the travel behaviours of food insecure populations, and our estimates might in fact be somewhat conservative if those households who do not report income rely more on walking and/or transit for their mobility needs.

>>>>

Food insecurity is a significant issue for many Canadian households and is associated with reductions in nutrition and poor mental and physical health outcomes. While emergency and community food services can provide some relief, the levels of need in the City of Hamilton were on the rise before the onset of the COVID-19 pandemic, with a troubling increase in the number of visits made by children to these services. While temporary financial relief measures may have helped in the summer of 2020, the pandemic has in all probability increased food insecurity for many households. To compound matters, the pandemic has also resulted in major disruptions, including to employment, mobility alternatives, and to emergency and community food services. In response, this research has sought to better understand accessibility to food banks and food service locations, as well as how the closure of some locations over the pre- and during-COVID time periods affected the potential for low-income households to reach these amenities. Using the balanced floating catchment area method to account for population demand and congestion effects at service points, we estimated multi-modal accessibility to emergency and community food service locations for low-income households. The weighting of travel time estimates by the modal split in different zonal geographies tailors the results to patterns of travel behaviour captured in the regional travel survey. Moreover, our parcel-level analysis presents a disaggregate approach to estimating accessibility based on the locations of residential parcels and dwellings.

Previous work has noted the important role of geography alongside other socio-economic and demographic indicators in household access to healthy food. The present papers is, to the best of our knowledge, among the first studies to focus on the geographic component of accessibility to emergency food services [@allen2021Changes]. Beyond accounting for the inflation of demand that occurs in traditional floating catchment area methods, our application of the balanced floating catchment area approach offers a novel analysis of accessibility to emergency and communal food services that is sensitive to the locations of low-income households, information on their age distributions and typical trip-making behaviour, the locations of services, their operations over time, and the characteristics of the city's multi-modal transportation network. Our results show that while accessibility levels were lower in the city's more car-oriented suburban and rural areas to begin with, the closure of 14% of the city's emergency and community food service locations during the pandemic resulted in an overall decrease in accessibility across the city. However, these effects were not uniform over space or for different population groups. Since the balanced floating catchment area method takes into account changes in demand and congestion for service providers, the closure of some services reverberates throughout the catchment areas of the whole city. For some suburban zones, the closure of a relatively high level-of-service location results in the remaining services being spread over a larger population. In others, high auto dependence for trips leads to decreases in accessibility that accumulate due to the loss of several locations initially reachable within 15 minutes by car. Reductions in accessibility in the city's more urban north end, where low-income households conduct higher proportions of trips by transit and walking, emphasize the importance of geographic proximity in the potential to reach service locations for these residents. Beyond geography, the results also highlight the differential impact of closures during COVID-19 on population groups, with seniors and children beign the two most impacted groups.

In the absence of information regarding how food insecure households travel to food banks and related services, we examined accessibility to food banks using a 15 minutes (one-way) travel time threshold. The fact that we must rely on a standard created for accessible drinking water in the developing world only serves to highlight the tragedy of food insecurity in an affluent country like Canada. More broadly, it points to the absurd need to understand how a bad situation was made worse by the pandemic: in effect, the analysis reveals that disparities in the need for emergency and community food services predated the pandemic, that the pandemic contributed to the deterioration of these services, and that populations already in distress, particularly children, ended up in an even more adverse state. How much worse, it is impossible to say, mainly because there is also a dearth of information, let alone standards, regarding _acceptable_ or _sufficient_ level of service when it comes to emergency food services. In fact, we would argue that the standard should be that no household faced food insecurity. As others have noted [e.g., @poppendieck1999sweet; @men2021food] the root of food insecurity is income poverty. There is likely a place for emergency food and community food services. In addition to providing food, these services could satisfy social needs by offering a social setting for seniors or by helping to connect households in need with longer term supports. From a food security perspective, on the other hand, these services should work only as a short term solution, and not as a semi-permanent feature of life for some of our fellow human beings. From a human rights perspective, long-term reliance on emergency food services should be as unacceptable in Canada as lack of clean drinking water within 30 minutes is elsewhere. Thus, while our analysis is valuable to map the suffering caused by food insecurity, from a policy perspective maintaining a robust social safety net that includes Employment Insurance and paid sick days are better tools to reduce this suffering than increasing the accessibility of emergency food services for food insecure populations.

In terms of future research, it is important to note that the degree to which our low-income cut-off of CAD\$40,000 reflects food insecurity in the different zones of the study area is not known. We also consider all emergency and communal food service locations equally. Information on differential capacities at food provider locations is currently not collected, and given the closure of facilities was not possible to obtain. Data on services, such as the number of meals served, could be used to refine the analysis in the future. Moreover, while the travel survey allows us to model multi-modal accessibilities that align with travel behaviour observed in the travel survey and capture differences in accessibility by age categories, the use of travel survey data for modelling food insecurity also has its limitations. Research into the population weighting methods used in the TTS note that the survey may under-count the lowest- and highest-income households in the survey study region, although the magnitude of this under-counting is unknown, and approximately 20% of respondents to the 2016 survey did not report their income [@rose2018transportation]. In that regard, the modal splits of low-income households observed through the TTS data may not accurately reflect the travel behaviours of food insecure populations, and our estimates might in fact be somewhat conservative if those households who do not report income rely more on walking and/or transit for their mobility needs.
